{\bf {\Large Filtering time series of dissolved oxygen for improved estimates of estuary metabolism}}

### 2014-09-29 abstract.tex
mcm revisions v2, still need to revise intro

In aquatic ecosystems, time series of \ac{DO} can be used to infer integrated ecosystem processes such as primary production, respiration, and net metabolism.  However, continuous monitoring data at estuaries may reflect variation from both biological and physical processes, such that observed data may produce inaccurate or misleading process estimates.  Statistical techniques that dynamically quantify variation in \ac{DO} and tidal changes over time have the potential to isolate biological signals in \ac{DO} variation.  A weighted regression method was developed to filter, or detide, the predicted \ac{DO} signal to remove the influence of physical advection on ecosystem metabolism estimates.  The method was tested using a simulation approach to create multiple \ac{DO} time series with known additive components of biological and physical variation on different periods.  The method was further validated using one year of continuous monitoring data at four water quality stations that are part of the \acl{NERRS}. We provide a detailed discussion on use of the method for improving certainty in evaluation of \ac{DO} measurements from sites with strong tidal influences.  Moreover, we propose that the method could expand use of the open-water method for estimating ecosystem metabolism in estuaries given that the approach can produce robust estimates of \ac{DO} that are independent of tidal advection.  In particular, this could facilitate the use of shorter deployment periods for water quality monitors or incomplete time series given that known biases related to water movement could be removed.

>>>>

In aquatic ecosystems, time series of \ac{DO} can be used to infer integrated ecosystem processes such as primary production, respiration, and net metabolism.  However, \ac{DO} time series data at estuaries may reflect variation from both biological and physical processes, potentially leading to inaccurate or misleading ecosystem process estimates.  One such physical process is the occurrence of large lateral \ac{DO} gradients in an estuary which may advect water with different \ac{DO} characteristics past a sensor.  In such situations, the lateral gradient may cause variation in \ac{DO} time series that are not attributable to metabolic processes.  Statistical techniques that dynamically quantify variation in \ac{DO} and tidal changes over time have the potential to isolate biological signals in \ac{DO} variation.  A weighted regression method was developed to filter the \ac{DO} time series to remove the influence of physical advection, thereby removing bias or noise in ecosystem metabolism estimates.  The method was tested using simulated \ac{DO} time series with known additive components of biological and physical variation.  The method was validated using one year of continuous monitoring data at four water quality stations that are part of the \acl{NERRS}. We provide a detailed discussion on use of the method for improving certainty in ecosystem metabolism estimates from sites with strong tidal influences.  This approach could improve metabolism estimates using shorter deployment periods or incomplete time series by removing bias attributed to lateral water movement.

### 2014-09-29 assess.Rnw
mcm revisions v2, still need to revise intro

Continuous monitoring data from the \acl{NERRS} was used to validate the weighted regression model by evaluating estimates of ecosytem metabolism obtained from observed and filtered \ac{DO} time series. \ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \aclu{SWMP} (\acs{SWMP}, \citetalias{CDMO14}).  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.  We selected sites from the \ac{SWMP} database that had desirable characteristics for validating weighted regression.  Specifically, four macrotidal sites were chosen based on apparent relationships between \ac{DO} and tidal changes (\cref{fig:case_map,tab:case_att}): Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).

@@

Half window widths of six days, one hour, and a tidal proportion of one half were used to filter the observed \ac{DO} time series.  Unlike the simulated data, the true biological \ac{DO} signal was unknown for the case studies.  Accordingly, the regression results were evaluated  using correlations of \ac{DO} and metabolism estimates with tidal height before and after application of the model.  Daily metabolism estimates before and after filtering were compared to the mean rate of tidal height change (i.e., first derivative of the predicted tidal height) for each day during separate solar periods.  Production rates were compared to mean rates of tidal height change during the day, respiration rates were compared to mean rates of change during the night, and net metabolism rates were compared to mean rates of change for the total 24 hour period each day.  Results were also evaluated based on the occurrence of `anomalous' daily production or respiration estimates, where anomalous was defined as negative production during the day and positive respiration estimates during the night.  Anomalous values have been previously attributed to the effects of physical processes on \ac{DO} time series \citep{Caffrey03}. Although anomalies could be caused by processes other than tidal advection, e.g., abiotic dark oxygen production \citep{Pamatmat97}, we assumed that physical processes were the dominant sources of these values given the tidal characteristics at each site.  Finally, means and standard errors of metabolism estimates were evaluated before and after filtering to determine if annual aggregations were significantly different.

>>>>

Results from the simulated time series were used to inform the validation of weighted regression with real data, specifically regarding half-window widths described below.  Continuous monitoring data from the \acl{NERRS} was used to validate the weighted regression model by evaluating estimates of ecosytem metabolism obtained from observed and filtered \ac{DO} time series. \ac{NERRS} is a federally-funded network of 28 protected estuaries established for long-term research, water-quality monitoring, education, and coastal stewardship \citep{Wenner04}.  Continuous water quality data have been collected at \ac{NERRS} sites since 1994 through the \aclu{SWMP} (\acs{SWMP}, \citetalias{CDMO14}).  In addition to providing a basis for trend evaluation, data from \ac{SWMP} provides an ideal opportunity to evaluate long-term variation in water quality parameters from biological and physical processes.  Continuous \ac{SWMP} data can be used to describe \ac{DO} variation at sites with different characteristics, including variation from ranges in tidal regime \citep{Sanger02} and rates of ecosystem production \citep{Caffrey03,Caffrey04}.  We selected sites from the \ac{SWMP} database that had desirable characteristics for validating weighted regression.  Specifically, four macrotidal sites were chosen based on apparent relationships between \ac{DO} and tidal changes (\cref{fig:case_map,tab:case_att}): Vierra Mouth station at Elkhorn Slough (California, 36.81$^{\circ}$N, 121.78$^{\circ}$W), Bayview Channel at Padilla Bay (Washington, 48.50$^{\circ}$N 122.50$^{\circ}$W), Middle Blackwater River station at Rookery Bay (Florida, 25.93$^{\circ}$N 81.60$^{\circ}$W), and Dean Creek station at Sapelo Island (Georgia, 31.39$^{\circ}$N 81.28$^{\circ}$W).

@@

Half window widths of six days, one hour, and a tidal proportion of one half were used to filter the observed \ac{DO} time series.  Although the selection of window widths involves a degree of subjectivity, results from the simulations suggested that these values were appropriate for filtering \ac{DO} times series within the constraints of the analysis. Unlike the simulated data, the true biological \ac{DO} signal was unknown for the case studies.  Accordingly, the regression results were evaluated  using correlations of \ac{DO} and metabolism estimates with tidal height before and after application of the model.  Daily metabolism estimates before and after filtering were compared to the mean rate of tidal height change (i.e., first derivative of the predicted tidal height) for each day during separate solar periods.  Production rates were compared to mean rates of tidal height change during the day, respiration rates were compared to mean rates of change during the night, and net metabolism rates were compared to mean rates of change for the total 24 hour period each day.  Results were also evaluated based on the occurrence of `anomalous' daily production or respiration estimates, where anomalous was defined as negative production during the day and positive respiration estimates during the night.  Anomalous values have been previously attributed to the effects of physical processes on \ac{DO} time series \citep{Caffrey03}. Although anomalies could be caused by processes other than tidal advection, e.g., abiotic dark oxygen production \citep{Pamatmat97}, we assumed that physical processes were the dominant sources of these values given the tidal characteristics at each site.  Finally, means and standard errors of metabolism estimates were evaluated before and after filtering to determine if annual aggregations were significantly different.

### 2014-09-29 intro.Rnw
mcm revisions v2, still need to revise intro

Time series of dissolved oxygen are increasingly used to estimate ecosystem metabolism \citep{Kemp12,Needoba12}.  Integrated measures of metabolism describe the balance between production and respiration processes that create and consume organic matter, respectively.  Although metabolic rates vary naturally between systems \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production \citep{Diaz08}.  Inputs of limiting nutrients beyond background concentrations may decrease the resilience of an ecosystem such that higher rates of production are coupled with higher biological oxygen demand \citep{Yin04,Kemp09}.  Cultural eutrophication is frequently linked to declines in water quality through lower levels of dissolved oxygen, degradation in aquatic vegetation habitat, and increased frequency of harmful algal blooms \citep{Cloern96,Short96,Rabalais02}.  Reliables estimates of ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

Open-water techniques have been used for decades to infer metabolic rates using \textit{in situ} measurements from continuous monitoring data \citep{Odum56}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  Originally conceived for streams \citep{Odum56}, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Hopkinson05,Caffrey13}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  Such assumptions are often only implicity verified in practice, leading to potential biases.  The fundamental assumption is that the time series of \ac{DO} describes the same water mass over time \citep{Needoba12}.  Estimates of metabolism from a single location may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Russell07}.  Application to lakes or estuaries have often assumed that a single sampling station provides sufficient data for estimating metabolism \cite{Staehr10}.  While single stations may be valid under specific conditions, numerous studies have shown that application of the open-water method to lakes or estuaries may be inappropriate given the effects of physical mixing \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}.

An extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  A significant portion of the estimates were negative for production and positive for respiration, i.e., opposite in sign given the assumptions of the method.  These `anamolous' values were attributed to the effects of tidal advection such that water masses with different metabolic histories were sampled at multiple sites. Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides. These studies provide compelling examples of the importance of potential bias in metabolism estimates related to physical advection.  The increasing availability of large, multi-annual datasets further warrants a need for quantiative methods that improve the accuracy of estimates of biological process rates.  \citet{Batt12} acknowledged this need by applying a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that account for cyclicity in time series associated with tidal variation in addition to process and observation uncertainty.

This article describes the application of a method for filtering an observed \ac{DO} time series for estimated tidal effects to more accurately quantify estimates of ecosystem metabolism for estuaries.  Specifically, the effects of tidal advection on \ac{DO} observations are removed to improve estimates of open-water metabolism with continuous water quality data.  We use a weighted regression approach developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  First, we use a simulation approach to create time series of \ac{DO} observations with known characteristics to evaluate ability of the weighted regression to remove the effects of tidal advection.  Second, four case studies were chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}) to illustrate the method for adjusting estimates of ecosystem metabolism.  Overall, the analysis is meant to better characterize the relative roles of biological and physical processes in estuarine systems.

>>>>

Time series of dissolved oxygen are increasingly used to estimate ecosystem metabolism \citep{Kemp12,Needoba12}.  Integrated measures of metabolism describe the balance between production and respiration processes that create and consume organic matter, respectively.  Although metabolic rates vary naturally at different spatial and temporal scales \citep{Ziegler98,Caffrey04,Russell07}, anthropogenic nutrient sources are often contributing factors that increase rates of production (\citealt{Nixon95}, \citetalias{NRC00}).  Inputs of limiting nutrients beyond background concentrations may decrease the resilience of an ecosystem such that higher rates of production are coupled with higher biological oxygen demand \citep{Yin04,Kemp09}.  Cultural eutrophication is frequently linked to declines in water quality through lower levels of dissolved oxygen, degradation in aquatic vegetation habitat, and increased frequency of harmful algal blooms \citep{Cloern96,Short96,Rabalais02,Diaz08}.  Reliables estimates of ecosystem metabolism are critical for measuring both background rates of production and potential impacts of human activities on ecosytem condition.

Open-water techniques have been used for decades to infer metabolic rates using \textit{in situ} measurements from continuous monitoring data \citep{Odum56}. Daily integrated measurements of metabolism represent the balance between daytime production and nighttime respiration.  The open-water method uses the diel fluctuation of dissolved oxygen to estimate ecosytem metabolism, after correcting for air-water gas exchange \citep{Kemp12}.  Originally conceived for streams \citep{Odum56}, the open-water method has been used with varying success in lakes \citep{Staehr10,Coloso11,Batt12} and estuaries \citep{Caffrey04,Russell07,Caffrey13}.  As with any method, the ability to accurately estimate whole system metabolism depends on the degree to which assumptions of the theory are met.  Such assumptions are often only implicity verified in practice, leading to potential biases.  The fundamental assumption is that the time series of \ac{DO} describes the same water mass over time \citep{Needoba12}.  Estimates of metabolism may be inaccurate if substantial variation in water column mixing occurs throughout the period of observation \citep{Russell07}.  Numerous studies have shown that application of the open-water method to lakes or estuaries may be problematic given the potential effects of physical mixing \citep{Ziegler98,Caffrey03,Coloso11,Batt12,Nidzieko14}.

An extensive analysis by \citet{Caffrey03} applied the open-water method to estimate metabolism at 28 continuous monitoring stations at 14 US estuaries.  An important observation from this study regarding the utility of the open-water method was that a significant portion of the estimates were negative (3 - 69\% depending on site). Significant variation from advection of water masses with different metabolic histories was suggested as a likely factor influencing the \ac{DO} time series. Inaccurate air-sea exchange estimates may have also contributed to these seemingly anomalous values.  Further, \citet{Nidzieko14} evaluated the effects of tidal advection on metabolism estimates in a mesotidal estuary.  Estimates were strongly correlated with the spring-neap cycle such that net heterotrophy was more common during spring tides, whereas metabolism was generally balanced during neap tides. These studies provide compelling examples of the importance of potential noise in metabolism estimates related to physical advection.  The increasing availability of large, multi-annual datasets further warrants a need for quantiative methods that improve the accuracy of estimates of biological process rates.  \citet{Batt12} acknowledged this need by applying a Kalman filter \citep{Harvey89} to remove process and observation uncertainty from \ac{DO} time series in lakes.  Similar approaches have not been developed for estuaries, particularly those that account for cyclicity in time series associated with tidal variation in addition to process and observation uncertainty.

This article describes the application of a method for filtering an observed \ac{DO} time series for estimated tidal effects to more accurately quantify estimates of ecosystem metabolism for estuaries.  Specifically, the apparent effects of tidal advection on \ac{DO} observations are removed to improve the fidelity of open-water metabolism estimates derived from continuous water quality data.  We used a weighted regression approach originally developed to resolve trends in pollutant concentrations in streams and rivers \citep{Hirsch10}.  The weighted regression approach creates dynamic predictions of \ac{DO} as a function of time and tidal height change, which are then used to filter, or detide, the \ac{DO} signal.  First, we used simulated \ac{DO} time series with known characteristics to evaluate ability of the weighted regression to remove the simulated effects of a tidally-advected \ac{DO} gradient.  Second, the simulation results informed the application of the method to four case studies chosen from the \aclu{NERRS} (\acs{NERRS}, \citealt{Wenner04}).  Specifically, one year of \ac{DO} time series for each case study was filtered to adjust estimates of ecosystem metabolism to apparent tidal effects.  In all examples, tidal height is used as a proxy for lateral water movements that may influence \ac{DO} observations.  In the absence of quantitative data describing lateral \ac{DO} variation (e.g., contemporaneous data along a tidal axis), we assume that tidal height is an appropriate and approximate measure that characterizes effects of physical advection.  Accordingly, `tidal variation' or `changes in tidal height' are used throughout in reference to assumed lateral \ac{DO} gradients that are carried past monitoring sensors by tidal currents.  Overall, the analysis is meant to better characterize the relative roles of biological and physical processes in estuarine systems.

### 2014-09-29 mat_proc.Rnw
mcm revisions v2, still need to revise intro

The weighted regression model for filtering \ac{DO} time series for tidal effects was adapted from the \ac{WRTDS} method \citep{Hirsch10}.  The functional form of the model is a simple linear regression:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where observed \ac{DO} is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each obervation with time as a proportion of the number of total observations added to each day.  The beginning of each day was considered the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period.  \ac{DO} variation was not modeled using this approach to avoid constraining parameter estimates by periodic, diel components.

@@

The normalization approach to filter tidal variation from the observed \ac{DO} time series differs slightly from methods in \citet{Hirsch10}.  The previous approach used a two-dimensional grid predicted for stream pollutant concentrations across the time series and the range of discharge values observed in the study system \citep{Hirsch10}.  Normalized or discharge-independent values for pollutant concentration were obtained by averaging grid predictions across the discharge values that were likely to occur on a given day.  Rather than creating a two-dimensional grid of \ac{DO} related to time and tidal height change, the normalized time series herein were the model predictions conditional on time and constant tidal height set to the mean:
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}
such that the normalized time series represents \ac{DO} variation related to biological processes.

>>>>

For this study, we adapted a weighted regression model to filter \ac{DO} time series for apparent tidal effects.  This model relied heavily on concepts used to develop the \ac{WRTDS} method for estimating pollutant concentrations in streams and rivers \citep{Hirsch10}.  The functional form of the model is:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $DO_{obs}$ is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each observation as a proportion of the number of total observations added to each day.  The beginning of each day was considered the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period.  \ac{DO} variation was not modeled using this approach to avoid constraining parameter estimates by periodic, diel components.

@@

The approach to filter physical advection from the observed \ac{DO} time series differs slightly from methods in \citet{Hirsch10}.  The previous approach used a two-dimensional grid predicted for stream pollutant concentrations across the time series and the range of discharge values observed in the study system \citep{Hirsch10}.  Normalized or discharge-independent values for pollutant concentration were obtained by averaging grid predictions across the discharge values that were likely to occur on a given day.  Rather than creating a two-dimensional grid of \ac{DO} related to time and tidal height change, the normalized time series herein were the model predictions conditional on time and constant tidal height set to the mean:
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}
such that the normalized time series represents \ac{DO} variation related to biological processes.  The term `filter' is used in reference to the removal of a specific variance component from the time series, while maintaining the structure of the biological component.  Although the approach shares similarities with common filtering techniques, a distinction is noted such that weighted regression has a specific purpose rather than more the generic objectives of common filters (e.g., moving window averages or local smoothers, \citealt{Shumway11}).

### 2014-09-29 mat_proc.tex
mcm revisions v2, still need to revise intro

\subsection{Weighted regression for modelling and detiding \ac{DO} time series}

The weighted regression model for detiding \ac{DO} time series was adapted from the \ac{WRTDS} method developed by \citet{Hirsch10}.  The \ac{WRTDS} method was developed to model pollutant concentration in streams and normalize predictions to changes in discharge.  The functional form of our model differs from the original model by relating observed \ac{DO} for to observation time and astronomical tidal height:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $t$ is decimal time and $H$ is tide height. Each observation contains a timestamp variable with date and time to the nearest half hour.  Decimal time  is calculated as a continuous value starting at zero and increasing with each time step.  Each unit of measurement (day, hour, etc.) is converted as a fraction of time on the annual scale and added to represent decimal time \citep{Hirsch10}.  The functional form differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period in addition to parameters $\beta_0$ to $\beta_2$.  Although \ac{DO} variation can follow a diel periodicity, \textit{in situ} measurements are poorly characterized by a sine wave.  For example, rates of change may be abrupt following diurnal variation in irradiance or daily \ac{DO} variation may be muted given the weather, as on cloudy days.  Sinuisoidal terms were not included in the model to avoid constraining the predictions to this assumption.

Weighted regression is similar to a moving window approach that allows for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. A separate regression model is estimated sequentially for each observation in the time series using a set of weights that are relevant to the point of reference (i.e., center of the window).  A single weight vector is calculated that quantifies the relevance of observations to the center of the window in respect to decimal time.  Specifically,  each observation is given a weight using a tri-cube weighting function \citep{Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ is inversely proportional to the distance $d$ from the center of the window.  Weights exceeding the maximum width of the window $h$ are equal to zero.  This approach gives higher importance to observations within the window that are relevant to the point of reference.  Rather than using different weight vectors for multiple variables as in \cite{Hirsch10}, a single weight vector was used for decimal time.  This weight vector implicitly accounted for variation in tidal height throughout the time series since the highest weights were in the center of the window, therefore giving highest importance to the tidal height occurring at the point of reference.

A nontrivial issue with weighted regression approach is the choice of window widths for calculating weights.  Excessively large or small window widths may respectively under- or over-fit the data.  Additionally, optimal window widths may depend on the objective for using the model.  The weighted regression approach can be used for both predicting \ac{DO} and normalizing to remove the variance in the \ac{DO} signal from tidal changes.  Optimal window widths that minimize prediction error or fit to the observed data are typically smaller than the optimum window widths for normalizing the time series.  Similarly, window widths that more effectively detide the \ac{DO} signal may produce predictions for the observed data that are not optimal.  Evaluations of the weighted regression method with simulated \ac{DO} time series described below used different window widths to identify an approximate optimal window width for detiding the \ac{DO} signal.  As such, the ability of the models to predict observed \ac{DO} was not a primary concern given that the optimal window width for detiding likely corresponds to a model that predicts \ac{DO} as a function of tide rather than observed \ac{DO} as a function of both tide and biological variation.

\subsection{Detiding the \ac{DO} signal using weighted regression}

The primary objective of the analysis was to evaluate ability of the weighted regression method to detide a \ac{DO} signal.  \citet{Hirsch10} developed the normalization approach for the \ac{WRTDS} method using a two-dimensional interpolation grid that contains predicted values of pollutant concentrations across the time series and the range of stream discharge values observed in the study system \citep{Hirsch10}.  Normalized values are obtained by averaging the predicted values across the range of discharge values that are likely to occur on a given day.  The normalized values represent variation in pollutant concentration that is independent of changes in discharge.

Predicted values of \ac{DO} concentration were normalized to remove variation from tidal height changes, although the approach herein differs from \citet{Hirsch10}.  Our adapted approach uses weighted regression to isolate sources of variation in the observed \ac{DO} signal that are related to unique effects of tidal height and biological process (\cref{fig:do_dtd}).  Two sets of values are predicted for the observed time series $DO_{obs}$, rather than creating an interpolation grid.  The first set of values uses the observed tidal height, whereas the second set uses the mean tidal height across the time series, $DO_{tid}$ and $DO_{mtd}$ respectively.  The first set of predictions represent \ac{DO} as a function of time and tide.  The second set of predictions represent the mean response of \ac{DO} conditional on time and a constant tidal height.  Both predictions are used to normalize or detide the \ac{DO} signal.  The residuals, $DO_{res}$ from the first set of predictions, i.e., the observed values minus the predicted values at changing tide, are considered to represent random variation in the \ac{DO} signal from biological processes independent of the tide and were added to $DO_{mtd}$ to create the final detided time series $DO_{dtd}$:
\begin{equation}
DO_{dtd} = DO_{mtd} + DO_{res}
\end{equation}
A critical assumption was that process and observation error in $DO_{res}$ are explicitly related to the biological \ac{DO} signal in addition to $DO_{mtd}$ from the second set of predicted values.

>>>>

\subsection{Weighted regression for modelling and filtering \ac{DO} time series}

For this study, we adapted a weighted regression model to filter \ac{DO} time series for apparent tidal effects.  This model relied heavily on concepts used to develop the \ac{WRTDS} method for estimating pollutant concentrations in streams and rivers \citep{Hirsch10}.  The functional form of the model is:
\begin{equation}\label{funform}
DO_{obs}= \beta_0 + \beta_1 t + \beta_2 H
\end{equation}
where $DO_{obs}$ is a linear function of time $t$ and tidal height $H$.  Time is a continuous variable for the day and time of each observation as a proportion of the number of total observations added to each day.  The beginning of each day was considered the nearest thirty minute observation to sunrise for the location.  Our model differed from the original \ac{WRTDS} method that included parameters to estimate variation of the response variable on a sinuisoidal period.  \ac{DO} variation was not modeled using this approach to avoid constraining parameter estimates by periodic, diel components.

Weighted regression was implemented as a moving window that allowed for estimation of \ac{DO} throughout the time series by adapting to variation through time as a function of tide. Regression models were estimated sequentially for each observation in the time series using dynamic weight vectors that change with the center of the window.  Weight vectors quantified the relevance of observations to the center of the window in respect to time, hour of the day, and tidal height.  Specifically, weights were assigned to each variable using a tri-cube weighting function \citep{Tukey77,Hirsch10}:
\begin{equation}
w= \left\{ 
  \begin{array}{l l}
    \left(1-\left(d/h\right)^3\right)^3 & \quad \textrm{if } |d| \leq h \\
    0 & \quad \textrm{if } |d| > h 
  \end{array} \right.
\end{equation}
where the weight $w$ of each observation is inversely proportional to the distance $d$ from the center of the window such that observations more similar to the point of reference are given higher importance in the regression.  Weights exceeding the maximum width of the window $h$ are equal to zero.  The tri-cube weighting function is similar to a Gaussian distribution such that weights decrease gradually from the center until the maximum window width is reached.  Regressions that use simpler windows (e.g., boxcar approach) are  more sensitive to influential observations as they enter or leave the window, whereas the tri-cube function minimizes their effect through gradual weighting of observations from the center \citep{Hirsch10}.  The final weight vector for each observation is the product of three separate weight vectors for time (day), hour, and tidal height. Windows for time and hour weight observations based on distance (time) from the center of the window.  The window for tidal height weights observations based on the difference from the center as a proportion of the total tidal height range.  For example, a half-window width of 0.5 means that observations are weighted proportionately within +/- 50\% the total range referenced to the tidal height in the center of the window. A low weight is given to an observation if any of the three weighting values were not similar to the center of the window since the final weight vector is the product of three weight vectors for each variable (see the link in the \hyperref[multi]{multimedia} section for graphical display of different weights).

The choice of window widths for weight vectors strongly affects the model results.  Excessively large or small window widths may respectively under- or over-fit the observed data.  Accordingly, appropriate window widths depend on the objective for using the model.  The weighted regression approach can be used for both predicting observed \ac{DO} and filtering the observed time series to remove the variance that coincided with the tidal cycle.  Window widths that minimize prediction error or fit to the observed data are typically smaller than widths that would be used for filtering tidal effects.  Similarly, window widths that more effectively filter the \ac{DO} signal may produce imprecise predictions for the observed data.  Evaluations of the weighted regression method with simulated \ac{DO} time series, described below, used multiple window widths to evaluate the ability of the model to filter the \ac{DO} signal.  The ability to predict observed \ac{DO} was not a primary objective such that the window widths were evaluated only in the context of removing tidal variation from the \ac{DO} time series.

The approach to filter physical advection from the observed \ac{DO} time series differs slightly from methods in \citet{Hirsch10}.  The previous approach used a two-dimensional grid predicted for stream pollutant concentrations across the time series and the range of discharge values observed in the study system \citep{Hirsch10}.  Normalized or discharge-independent values for pollutant concentration were obtained by averaging grid predictions across the discharge values that were likely to occur on a given day.  Rather than creating a two-dimensional grid of \ac{DO} related to time and tidal height change, the normalized time series herein were the model predictions conditional on time and constant tidal height set to the mean:
\begin{equation} \label{do_nrm}
DO_{nrm} = f(DO_{obs}|\bar{H}, t)
\end{equation}
such that the normalized time series represents \ac{DO} variation related to biological processes.  The term `filter' is used in reference to the removal of a specific variance component from the time series, while maintaining the structure of the biological component.  Although the approach shares similarities with common filtering techniques, a distinction is noted such that weighted regression has a specific purpose rather than more the generic objectives of common filters (e.g., moving window averages or local smoothers, \citealt{Shumway11}).
